True versus apparent shapes of astrophysical bow shocks

### 2017-02-09 papers/Paper1/sec-projection.tex
Incorporated A' and q' results in  parabola appendix.
Restored projection section.
Correct errors and footnote
Minor changes:
* Exchange Figures 2 and 3/
* subindices of eqautions A9 and A10 converted to \mathrm
Minor corrections to secs 3.3 and 3.4

%Note: I'm aware that some of this material should be moved to an appendix, but I think it will be a future edition.
\subsection{Frames of reference}

@@

@@

\subsection{Characteristic radii}
\label{sec:characteristic-radii}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/characteristic-radii}
  \caption{Characteristic radii in the body frame.}
  \label{fig:characteristic-radii}
\end{figure}

\subsection{Unit vectors normal and tangential to the surface}
\label{sec:unit-vectors-normal}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/bowshock-unit-vectors}
  \caption{Unit vectors in the body frame that are normal and
    tangential to the surface \(R(\theta)\) in a plane of constant
    azimuth, \(\phi\).}
  \label{fig:unitvec}
\end{figure}
We define unit vectors \(\uvec{n}, \uvec{t}\), such that \(\uvec{n}\) is normal to the surface, while \(\uvec{t}\) is tangent to the surface in a plane of constant \(\phi\). For \(\phi = 0\) the surface lies in the \(xy\) plane and it is straightforward to show (Fig.~\ref{fig:unitvec}) that in this case the unit vectors are given by 
\begin{equation}
  \label{eq:uvec-phi0}
  \uvec{t}_0 =
  \begin{Vector}
    -\cos\alpha \\ \sin\alpha \\ 0
  \end{Vector}
  \quad \mathrm{and} \quad
  \uvec{n}_0 =
  \begin{Vector}
    \sin\alpha \\ \cos\alpha \\ 0
  \end{Vector}, 
\end{equation}
where
\begin{equation}
  \label{eq:alpha}
  \tan\alpha = -\left.\frac{dy}{dx}\right\vert_{R(\theta)} 
  = \frac{1 + \omega \tan\theta}{\tan\theta - \omega}
\end{equation}
and 
\begin{equation}
  \label{eq:omega}
  \omega(\theta) = \frac{1}{R} \frac{dR}{d\theta} . 
\end{equation}
For general \(\phi\), \(\uvec{n}\) and \(\uvec{t}\) can be found by rotating equations~(\ref{eq:uvec-phi0}) around the \(x\)-axis, which can then be converted to the observer frame by use of equation~(\ref{eq:Trans}) to give
\begin{gather}
  \label{eq:nprime}
  \begin{split}
    \uvec{n}' = & \frac{1}{(1+\omega^2)^{1/2}} \\
    & \times 
      \begin{Vector}
        (\cos\theta+\omega\sin\theta)\cos i 
        - (\sin\theta-\omega\cos\theta) \sin i \sin\phi\\
        (\sin\theta-\omega\cos\theta)\cos\phi \\
        (\cos\theta+\omega\sin\theta)\sin i 
        + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
      \end{Vector}\\
  \end{split}\\
  \label{eq:tprime}
  \begin{split}
    \uvec{t}' = & \frac{1}{(1+\omega^2)^{1/2}} \\
    & \times 
      \begin{Vector}
        -(\sin\theta-\omega\cos\theta)\cos i
        - (\cos\theta+\omega\sin\theta)\sin\phi\sin i \\
        (\cos\theta+\omega\sin\theta)\cos\phi \\
        -(\cos\theta+\omega\sin\theta)\sin i 
        + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
      \end{Vector}
  \end{split}
  \end{gather}

\subsection{Tangent line}

@@

The boundary on the plane of the sky of the projected surface is the
locus of those lines of sight that graze the surface tangentially.
This corresponds to a curved line on the surface itself, which we
denote the \textit{tangent line}, and which is defined by the
condition 
\begin{equation}
  \label{eq:tangent-line-condition}
  \uvec{n}'\cdot \uvec{z}' = 0.
\end{equation}
We denote by \(\phi\T\) that value of $\phi$ that satisfies this
relation for a given inclination, \(i\), and polar angle, \(\theta\):
\begin{equation}
\sin\phi\T = \tan i\tan\alpha = \tan i \frac{1+\omega\tan\theta}{\omega-\tan\theta}
\label{eq:tanphi}
\end{equation}
From equations~(\ref{eq:body-frame}, \ref{eq:Trans}) it follows that
the observer-frame coordinates of the tangent line are given by
\begin{equation}
\left(\begin{array}{c}
x'\T \\ y'\T \\ z'\T
\end{array}\right)= R(\theta)\left(\begin{array}{c}
\cos\theta\cos i - \sin\theta\sin\phi\T \sin i \\
\sin\theta(1-\sin^2\phi\T)^{1/2} \\
\cos\theta\sin i +\sin\theta\sin\phi\T\cos i
\end{array}\right).
\label{eq:tangential}
\end{equation} 
Note that, in general, \(z'\T\) is not a linear function of \(x'\T\)
and \(y'\T\), so that the tangent line is not a plane curve.

The apparent shape \((x'\T, y'\T)\) of the tangent line on the plane
of the sky can also be described in polar form as \(R'(\theta')\),
where
\begin{equation}
  \label{eq:R-prime-theta-prime}
  R' = (x'^2\T+ y'^2\T)^{1/2} 
  \quad \text{and} \quad
  \tan\theta' = y'\T / x'\T.
\end{equation}
It is important to note that equation~(\ref{eq:tanphi}) does not have
a solution for arbitrary values of $\theta$ and $i$, but only when
$|\tan i\tan\alpha|<1$. In particular, if $i\neq 0$, then the tangent
line only exists for \(\theta > \theta_{0}\) where \(\theta_{0}\) is
given implicitly by
\begin{align}
\tan\theta_{0} = \frac{|\tan i| + \omega(\theta_{0})}{1-\omega(\theta_{0}) |\tan i|} . 
\label{eq:thetapar}
\end{align}
In addition, if the surface is sufficiently ``open''
(\(\alpha > \alpha_{\mathrm{min}} > 0\) for all \(\theta\)), then for
those inclinations with
\(\vert i\vert > (90^\circ - \alpha_{\mathrm{min}}) \) the tangent
line does not exist for any value of \(\theta\).  In other words, when
the viewing angle is sufficiently close to face-on, the projected
surface has no ``edge'' and will no longer appear to be a bow shock.

@@

Considering further applications to bow shocks, we will consider open shells. In order to compare the shell shape given by $R(\theta)$ with observations,
it is convenient to define the following apparent radii in the observer frame: $R'_{0}$ and $R'_{90}$. These are projected distances of the shell tangent line
from the origin. The first is measured in the direction of the symetry axis, and the second in a perpendicular direction. More concretely $R'_{0} = x'\T(y'\T=0)$
and $R'_{90} = y'\T(x'\T=0)$. From equations (\ref{eq:tanphi}) and (\ref{eq:tangential}) we find that:
\begin{align}
R'_{0} = R(\theta_{0})\cos(\theta + i) \label{eq:Rpar} 
\end{align}
Where $\theta_{0}$ is the solution of equation (\ref{eq:thetapar}), and
\begin{align}
R'_{90} = R(\theta_{90})\sin\theta_{90}\left(1-\sin^2(\phi\T(\theta_{90}))\right)^{1/2}
\end{align}
Where $\theta_{90}$ is the solution of the next implicit equation:
\begin{align}
\cot\theta_{90} = \frac{1-\left(1+\omega(\theta_{90})^2\sin^22i\right)^{1/2}}{2\omega(\theta_{90})\cos^2 i}
\end{align}

>>>>

% Note: I'm aware that some of this material should be moved to an appendix, but I think it will be a future edition.

\subsection{Frames of reference}
\label{sec:ref-frames}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/characteristic-radii}
  \caption{Characteristic radii in the body frame.}
  \label{fig:characteristic-radii}
\end{figure}

@@

\subsection{Unit vectors normal and tangential to the surface}
\label{sec:unit-vectors-normal}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/bowshock-unit-vectors}
  \caption{Unit vectors in the body frame that are normal and
    tangential to the surface \(R(\theta)\) in a plane of constant
    azimuth, \(\phi\).}
  \label{fig:unitvec}
\end{figure}
We define unit vectors \(\uvec{n}, \uvec{t}\), such that \(\uvec{n}\) is normal to the surface, while \(\uvec{t}\) is tangent to the surface in a plane of constant \(\phi\). For \(\phi = 0\) the surface lies in the \(xy\) plane and it is straightforward to show (Fig.~\ref{fig:unitvec}) that in this case the unit vectors are given by 
\begin{equation}
  \label{eq:uvec-phi0}
  \uvec{t}_0 =
  \begin{Vector}
    -\cos\alpha \\ \sin\alpha \\ 0
  \end{Vector}
  \quad \mathrm{and} \quad
  \uvec{n}_0 =
  \begin{Vector}
    \sin\alpha \\ \cos\alpha \\ 0
  \end{Vector}, 
\end{equation}
where
\begin{equation}
  \label{eq:alpha}
  \tan\alpha = -\left.\frac{dy}{dx}\right\vert_{R(\theta)} 
  = \frac{1 + \omega \tan\theta}{\tan\theta - \omega}
\end{equation}
and 
\begin{equation}
  \label{eq:omega}
  \omega(\theta) = \frac{1}{R} \frac{dR}{d\theta} . 
\end{equation}
For general \(\phi\), \(\uvec{n}\) and \(\uvec{t}\) can be found by rotating equations~(\ref{eq:uvec-phi0}) around the \(x\)-axis, which can then be converted to the observer frame by use of equation~(\ref{eq:Trans}) to give
\begin{gather}
  \label{eq:nprime}
  \begin{split}
    \uvec{n}' = & \frac{1}{(1+\omega^2)^{1/2}} \\
    & \times 
      \begin{Vector}
        (\cos\theta+\omega\sin\theta)\cos i 
        - (\sin\theta-\omega\cos\theta) \sin i \sin\phi\\
        (\sin\theta-\omega\cos\theta)\cos\phi \\
        (\cos\theta+\omega\sin\theta)\sin i 
        + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
      \end{Vector}\\
  \end{split}\\
  \label{eq:tprime}
  \begin{split}
    \uvec{t}' = & \frac{1}{(1+\omega^2)^{1/2}} \\
    & \times 
      \begin{Vector}
        -(\sin\theta-\omega\cos\theta)\cos i
        - (\cos\theta+\omega\sin\theta)\sin\phi\sin i \\
        (\cos\theta+\omega\sin\theta)\cos\phi \\
        -(\cos\theta+\omega\sin\theta)\sin i 
        + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
      \end{Vector}
  \end{split}
  \end{gather}
\subsection{Tangent line}

@@

The boundary on the plane of the sky of the projected surface is the
locus of those lines of sight that graze the surface tangentially.
This corresponds to a curved line on the surface itself, which we
denote the \textit{tangent line}, and which is defined by the
condition  
\begin{equation}
  \label{eq:tangent-line-condition}
  \uvec{n}'\cdot \uvec{z}' = 0.
\end{equation}

We denote by \(\phi\T\) that value of $\phi$ that satisfies this
relation for a given inclination, \(i\), and polar angle, \(\theta\):
\begin{equation}
\sin\phi\T = \tan i\tan\alpha = \tan i \frac{1+\omega\tan\theta}{\omega-\tan\theta}
\label{eq:tanphi}
\end{equation}
From equations~(\ref{eq:body-frame}, \ref{eq:Trans}) it follows that
the observer-frame coordinates of the tangent line are given by
\begin{equation}
\left(\begin{array}{c}
x'\T \\ y'\T \\ z'\T
\end{array}\right)= R(\theta)\left(\begin{array}{c}
\cos\theta\cos i - \sin\theta\sin\phi\T \sin i \\
\sin\theta(1-\sin^2\phi\T)^{1/2} \\
\cos\theta\sin i +\sin\theta\sin\phi\T\cos i
\end{array}\right).
\label{eq:tangential}
\end{equation} 
Note that, in general, \(z'\T\) is not a linear function of \(x'\T\)
and \(y'\T\), so that the tangent line is not a plane curve.

The apparent shape \((x'\T, y'\T)\) of the tangent line on the plane
of the sky can also be described in polar form as \(R'(\theta')\),
where
\begin{equation}
  \label{eq:R-prime-theta-prime}
  R' = (x'^2\T+ y'^2\T)^{1/2} 
  \quad \text{and} \quad
  \tan\theta' = y'\T / x'\T.
\end{equation}
Note that equation~(\ref{eq:tanphi}) will not have
a solution for arbitrary values of $\theta$ and $i$, but only when
$|\tan i\tan\alpha|<1$. In particular, if $i\neq 0$, then the tangent
line only exists for \(\theta > \theta_{0}\) where \(\theta_{0}\) is the value of \(\theta\) on the tangent line's projected symmetry axis (\(\theta' = 0\)), 
given implicitly by
\begin{align}
\tan\theta_{0} = \frac{|\tan i| + \omega(\theta_{0})}{1-\omega(\theta_{0}) |\tan i|} . 
\label{eq:thetapar}
\end{align}
In addition, if the surface is sufficiently ``open''
(\(\alpha \ge \alpha_{\mathrm{min}} > 0\) for all \(\theta\)), then for those
inclinations with
\(\vert i\vert > (90^\circ - \alpha_{\mathrm{min}}) \) the tangent line does not exist
for any value of \(\theta\).  In other words, when the viewing angle is
sufficiently close to face-on, the projected surface has no ``edge''
and will no longer look like a bow shock to the observer.

@@

% Considering further applications to bow shocks, we will consider open shells.
In order to compare the shell shape given by $R(\theta)$ with observations,
it is convenient to define the following apparent radii in the observer frame: $R'_{0}$ and $R'_{90}$. These are projected distances of the shell tangent line
from the origin. The first is measured in the direction of the symmetry axis, and the second in a perpendicular direction. More concretely $R'_{0} = x'\T(y'\T=0)$
and $R'_{90} = y'\T(x'\T=0)$. From equations (\ref{eq:tanphi}) and (\ref{eq:tangential}) we find that:
\begin{align}
R'_{0} = R(\theta_{0})\cos(\theta_{0} + i) \label{eq:Rpar} 
\end{align}
Where $\theta_{0}$ is the solution of equation (\ref{eq:thetapar}), and
\begin{align}
R'_{90} = R(\theta_{90})\sin\theta_{90}\left(1-\sin^2\bigl(\phi\T(\theta_{90})\bigr)\right)^{1/2}
\end{align}
Where $\theta_{90}$ is the solution of the next implicit equation:
\begin{align}
\cot\theta_{90} = \frac{1-\left(1+\omega(\theta_{90})^2\sin^22i\right)^{1/2}}{2\omega(\theta_{90})\cos^2 i}
\end{align}
